Active Learning from Blackbox to Timed Connectors

### 2016-01-19 paper/paper.tex
.

% finally no todos should exist in this draft
% \usepackage[textsize=tiny]{todonotes}

@@

\begin{document}
\maketitle 
\begin{abstract}
  Coordination models and languages play a key role in formally specifying the communication and
  interaction among different components in large-scale distributed and concurrent systems. In this
  paper, we propose an active learning framework to extract timed connector models from black-box
  system implementation. 
  We first introduce parameterized Mealy machine(PMM) as an operational semantic
  model for channel-based coordination language Reo. Parameterized Mealy machine serves as a bridge
  between Reo connectors and mealy machines. With the product operator, complex connectors can be
  constructed by joining basic channels and transformed into mealy machines. Moreover, we adapt L*,
  a well-known learning algorithm, to timed connectors (in the form of mealy machines). The new
  algorithm has shown its efficiency in multiple case studies. 
  Implementations of this framework is provided as a package in \texttt{Golang}.
\end{abstract}

@@

Distributed real-time embedded systems (DRES) are reforming our lives with the
Internet of Things(IoT), wherein individual components are composed via a
\emph{connector} to build a system. Such systems could be distributed logically or physically, which
makes coordination processes even more complicated. In this case, we need to specify the coordination
processes with \emph{coordination languages}, such that formal techniques can be applied to
guarantee their reliability.

\emph{Timed Reo} is a real-time extension of the coordination language \emph{Reo}, by which
coordination process in DRES can be depicted clearly and intuitively. Different formal
semantics are proposed to specify the behavior of timed Reo.
For example, in \cite{DBLP:conf/tase/Meng12}, a UTP-based
(\emph{Unifying Theories of Programming}) semantics is provided to verify connectors as
\emph{designs}. An operational semantics based on \emph{constraint automata} was raised by Baier et
al.\cite{DBLP:conf/sefm/ArbabBBR04}, where a variant of LTL was also proposed to describe the properties of timed connectors.

Formal verification and validation techniques have also been proved applicable for timed connectors, 
e.g. a comformance testing framework on timed connectors in \cite{DBLP:conf/tase/LiCWS15}.
A SAT-based bounded model checking approach was adapted in \cite{DBLP:journals/scp/Kemper12} for
verifying connectors. Although these solutions are practical and impressive, there is one important
question faced by most of them: \emph{how to obtain the formal models?}

\emph{Correctness} of connectors is closely related to some low-level implementation details.
For example, well-written code may behave dramatically weird with an improper set of concurrency
primitives. Such senarios happen frequently in embedded systems with different hardware platforms or
operating systems. Consequently, manually modeling an existing connector seems rather unreliable,
even with reference to its source code.

As a branch of \emph{machine learning} technique, active learning offers a way to obtain models from
low-level models. Works in \cite{DBLP:journals/mt/Daelemans10, DBLP:journals/iandc/Angluin87,
DBLP:conf/fase/RaffeltS06} shows interesting examples where active learning is used to extract
\emph{Mealy machines} or \emph{regular languages} without time domain.

@@

In this paper we address this question by proposing a active learning framework as shown in Figure
\ref{fig:howto}) to automatically extract timed connector models from blackbox implementations. 
To achieve this goal, we first introduce \emph{parameterized Mealy machine} as a parametierized
semantics for timed Reo connectors, which can be transformed intoe concrete Mealy machines with a
given alphabet. Then the $L*$ algorithm \cite{DBLP:journals/iandc/Angluin87} is adapted and
optimized to extract Mealy machines with timed action from the connector implementations as
blackboxes.

The rest of the paper is organized as follows: In Section
\ref{sec:preliminaries} we briefly illustrate some basic concepts, including the coordination
language Reo, Mealy machine, and active automata learning. Section \ref{sec:semantics} defines an
operational semantics of Reo, which is used in Section \ref{sec:activelearning} to show how to
extract Reo models from blackboxes by means of active learning. In Section
\ref{sec:experiment}, we discuss the implementation and optimization of the approach in
\texttt{Golang}.

@@

\begin{itemize}
  \item [-] A \emph{Sync} channel accepts a data item from its source end iff the data item can be
    dispensed through its sink end simultaneously.
  \item [-] A \emph{LossySync} channel is always ready to accept data items. These items will be
    dispensed through its sink end if possible, otherwise they will be lost.
  \item [-] A \emph{SyncDrain} channel has two source ends and no sink end. It can accept a data
    item through one of its source end iff a data item is also available to be accepted
    simultaneously through the other end. Then both items will be dropped.
  \item [-] A \emph{FIFO1} channel is an asychronous channel with a buffer cell. It accepts a
    data item whenever the buffer is empty. 
  \item [-] A t-timer channel accepts any data item from its source end, and later dispense it to
    its sink end after a delay of $t$ time units.

\end{itemize}

@@

\begin{itemize}
  \item[-] \textbf{Membership Query} (hereinafter referred to as \emph{mq}) In grammar-learning
    \cite{DBLP:journals/iandc/Angluin87}) \emph{mq} checks if a word is a member of certain language
    defined by the given grammar. When it comes to
    automata learning, \emph{mq} is supposed to provide \emph{simulation results} for given input
    sequences.
  \item[-] \textbf{Equivalence Query} (hereinafter referred to as \emph{eq}) Given a hypothesis
    (usually constructed by the learning algorithm), \emph{eq} checks whether the hypothesis is
    equivalent to the system-under-learn and generates a counter-example if needed. Generally, 
    \emph{eq} is irrealizable when SUL is a blackbox. So we tactically use $mq$ to achieve the
    approximate results.
\end{itemize}

@@

\subsection{Time Domain}
Time has been investigated in several extensions of Reo. For example, timed
Reo\cite{DBLP:conf/sefm/ArbabBBR04, DBLP:conf/fmoods/MengA07}, hybrid Reo\cite{DBLP:conf/icfem/ChenSS14}, etc.
Generally, these models are designed to handle real-time behavior where time is defined on the
real number field $\mathbb{R}$. However, there are also found some works like
\cite{DBLP:journals/fmsd/PrabhakarDM015} where
rational time indeed  makes things easier. In this paper, we choose the rational number field
$\mathbb{Q}$ as our time domain, which simplifies discretization of timed behavior greatly.

As presented in section \ref{sec:reo}, all real-time behavior in timed Reo comes with a finite set
of timed channels, and the number of these channels are apparently finite. We use
$t_i\in\mathbb{Q}, i=1,2,\dots,n$ to denote the delays of these timed channels, and then we can define a precision
function $prec$ which calculate a gcd-style \emph{maximal time precision}:
\[
prec(t_1,\dots,t_n) = \max\{T\in\mathbb{Q}|\forall t_i.\exists n_i\in\mathbb{N}.t_i=n_i\cdot T\}
\]
It's easy to prove that such a $T$ always exists.

In embedded systems, the maximal time precision is also called ``clock-period'', which is the basic
time unit provided by an oscillator. For example, a widely-used Intel-80C51 microcontroller works
under the frequency 16Mhz, where the clock-period is $62.5$ns. A resonable assumption comes that
such a maximal time precision $dt=prec(t_1,\dots,t_n)$ should always be provided to make sure
different components are able to work together. Then all $t_i$-Timers can be seen as $n_idt$-Timers
for some $n_i$. In the following sections, we'll use $n_i$-Timers instead.

@@

\subsection{Parameterized Mealy Machine}
We present a model named \emph{parameterized Mealy machine} to model timed connectors. PMM is
supposed to behave as a bridge between Mealy machines and timed connectors. Connectors are firstly
defined as PMMs, and composed by production operator and link operator. Then, concrete Mealy-machine
model can be generated from the PMM model.
Following the formal definition of Mealy machine in Section \ref{sec:preliminaries}, we define the
\emph{parameterized Mealy machine} as follows.

@@

\begin{definition}[Production]
  The production operator \emph{prod} of two parameterized Mealy machines
  \[
  \mathcal{PM'} = prod(\mathcal{PM}_1,\mathcal{PM}_2)\mbox{ if }\mathcal{PM}_2.O\cap
  \mathcal{PM}_1.I=\varnothing
  \]
  is defined as follows.
  \begin{itemize}
  	\item[-] $\forall\Sigma. \mathcal{PM}'.S(\Sigma)=\mathcal{PM}_1.S(\Sigma)\times \mathcal{PM}_2.S(\Sigma)$
    \item[-] $\mathcal{PM}'.I=(\mathcal{PM}_1.I\cup \mathcal{PM}_2.I)\backslash \mathcal{PM}_1.O$
    \item[-] $\mathcal{PM}'.O=(\mathcal{PM}_1.O\cup \mathcal{PM}_2.O)\backslash \mathcal{PM}_2.I$
  \end{itemize}
  \emph{(Here we assume that sink ends of $\mathcal{PM}_1$ can be connected to source ends $\mathcal{PM}_2$, but not
  vise versa)}
  \begin{itemize}
    \item[-] $\mathcal{PM}'.s_0=(\mathcal{PM}_1.s_0, \mathcal{PM}_2.s_0)$
    \item[-] $\forall\Sigma. \mathcal{PM}'.\delta(\Sigma)((s_1,s_2), i)=$
      \begin{displaymath}
        \left\{
        \begin{array}[ht]{ll}
          (Out_1 \cup Out_2)\downarrow_{\mathcal{PM}'.O} & Out_1\neq\rblock\land Out_2\neq\rblock \\
          \rempty & i = T \\
          \rblock & otherwise \\
        \end{array}
        \right.
      \end{displaymath}
      where we have
      \begin{itemize}
        \item[*] $In_1 = i\downarrow_{\mathcal{PM}_1.I}$
        \item[*] $Out_1 = \mathcal{PM}_1.\delta(\Sigma)(s_1,In_1)$
        \item[*] $In_2 = (Out_1 \cup i)\downarrow_{\mathcal{PM}_2.I}$
        \item[*] $Out_2 = \mathcal{PM}_2.\delta(\Sigma)(s_2,In_2)$
      \end{itemize}
    \item[-] $\forall\Sigma. \mathcal{PM}'. \lambda(\Sigma)((s_1,s_2),i)=(s_1',s_2')$
      where we have
      \begin{itemize}
        \item[*] $s_1' = \mathcal{PM}_1.\lambda(\Sigma)(s_1,In_1)$
        \item[*] $s_2' = \mathcal{PM}_2.\lambda(\Sigma)(s_2,In_2)$
      \end{itemize}
  \end{itemize}
\end{definition}

The idea in \emph{production} is quite simple. The output of $\mathcal{PM}_1$ would be provided as part of the
input of $\mathcal{PM}_2$. Time actions T is only executed simutaneously between $\mathcal{PM}_1$ and $\mathcal{PM}_2$.

@@

\subsection{Observation Table}
\emph{Observation Tables}, proposed in \cite{DBLP:journals/iandc/Angluin87}, is used to construct
a hypothetical Mealy machine from results membership queries.

@@

@@

Our future works would mainly focus on better support of dense time. To describe dense
time behavior, the Mealy machine model needs a lot of changes instead of a simple T action. Besides,
we will also try to improve our algorithm to handle non-deterministic behaviors. This is not a brand
new topic \cite{DBLP:journals/eceasst/VolpatoT15}, but we believe that there is still room for
improvement.

\bibliographystyle{IEEEtran}
% \bibliography{IEEEabrv, bib}
\bibliography{bib}

>>>>

% finally no todos should exist in this draft
\usepackage[textsize=tiny]{todonotes}

@@

\begin{document}
\maketitle 
\begin{abstract}
  Coordination models and languages play a key role in formally specifying the communication and
  interaction among different components in large-scale distributed and concurrent systems. In this
  paper, we propose an active learning framework to extract timed connector models from black-box
  system implementation. 
  We first introduce parameterized Mealy machine(PMM) as an operational semantic
  model for channel-based coordination language Reo. PMM serves as a bridge
  between Reo connectors and Mealy machines. With the product operator, complex connectors can be
  constructed by joining basic channels and the PMM of connectors can be transformed into Mealy
  machines. Moreover, we adapt L*, a well-known learning algorithm, to timed connectors (in the form
  of Mealy machines). The new algorithm has shown its efficiency in multiple case studies. 
  This framework has been implemented as a package in \texttt{Golang}.
\end{abstract}

@@

Distributed real-time embedded systems (DRES) are reforming our lives with the
Internet of Things(IoT), wherein individual components are composed via \emph{connectors} 
to build systems. Such systems could be distributed logically or physically, which
makes coordination processes even more complicated. In this case, we need to specify the coordination
processes with \emph{coordination languages}, such that formal techniques can be applied to
guarantee their reliability.

\emph{Timed Reo} is a real-time extension of the coordination language \emph{Reo}, which can be used
to describe the coordination process in DRES clearly and intuitively. Different formal
semantics have been proposed to specify the behavior of timed Reo.
For example, an operational semantics based on \emph{timed constraint automata} was raised by Baier et
al.\cite{DBLP:conf/sefm/ArbabBBR04}, where a variant of LTL was also proposed to describe the
properties of timed connectors.
In \cite{DBLP:conf/tase/Meng12}, a UTP design model was proposed to specify the behavior of timed
connectors..

Formal verification and validation techniques have also been proved applicable for timed connectors, 
e.g. a comformance testing framework on timed connectors has been proposed in \cite{DBLP:conf/tase/LiCWS15}.
A SAT-based bounded model checking approach was adapted in \cite{DBLP:journals/scp/Kemper12} for
verifying connectors. Although these solutions are practical and impressive, there is one important
question faced by most of them: \emph{how to obtain the formal models?}

\emph{Correctness} of connectors is closely related to some low-level implementation details.
For example, well-written code may behave dramatically weird with an improper set of concurrency
primitives. Such scenarios happen frequently in embedded systems with different hardware platforms or
operating systems. Consequently, manually extracting a proper model for an existing connector
implementation seems rather unreliable, even with reference to its source code.

As a branch of \emph{machine learning} technique, active learning offers a means to obtain
high-level models from low-level models. Works in \cite{de2010grammatical,
DBLP:journals/iandc/Angluin87, DBLP:conf/fase/RaffeltS06} show interesting examples where active
learning is used to extract \emph{Mealy machines} or \emph{regular languages} without time domain.

@@

In this paper we address this question by proposing a active learning framework as shown in Figure
\ref{fig:howto} to automatically extract timed connector models from blackbox implementations. 
To achieve this goal, we first introduce \emph{parameterized Mealy machine} as a parameterized
semantics for timed Reo connectors, which can be transformed to concrete Mealy machines with a
given alphabet. Then the $L*$ algorithm \cite{DBLP:journals/iandc/Angluin87} is adapted and
optimized to extract Mealy machines with timed action from the connector implementations as
blackboxes.

The rest of the paper is organized as follows: In Section
\ref{sec:preliminaries} we briefly illustrate some basic concepts, including the coordination
language Reo, Mealy machine, and active automata learning. Section \ref{sec:semantics} defines an
operational semantics of Reo, which is used in Section \ref{sec:activelearning} to show how to
extract Reo models from blackboxes by means of active learning. In Section
\ref{sec:experiment}, we discuss the implementation and optimization of the approach in
\texttt{Golang}. Finally, Section VI concludes the paper.

@@

\begin{itemize}
  \item [-] A \emph{Sync} channel accepts a data item from its source end iff the data item can be
    dispensed through its sink end simultaneously.
  \item [-] A \emph{LossySync} channel is always ready to accept data items. These items will be
    dispensed through its sink end simultaneously if possible, otherwise they will be lost.
  \item [-] A \emph{SyncDrain} channel has two source ends and no sink end. It can accept a data
    item through one of its source end iff a data item is also available to be accepted
    simultaneously through the other end. Then both items will be dropped.
  \item [-] A \emph{FIFO1} channel is an asychronous channel with a buffer cell. It accepts a
    data item whenever the buffer is empty. 
  \item [-] A t-timer channel accepts any data item from its source end, and later dispense it to
    its sink end after a delay of $t$ time units.
    \footnote{Here the interpretation of t-Timer is a bit different as in other works like
    \cite{DBLP:conf/sefm/ArbabBBR04, DBLP:conf/tase/Meng12}, but we can easily transform between the
    two kinds of behavior by composing with some other basic channels.}
\end{itemize}

@@

\begin{itemize}
  \item[-] \textbf{Membership Query} (hereinafter referred to as \emph{mq}) In grammar-learning
    \cite{DBLP:journals/iandc/Angluin87}, \emph{mq} checks if a word is a member of certain language
    defined by the given grammar. When it comes to
    automata learning, \emph{mq} is supposed to provide \emph{simulation results} for given input
    sequences.
  \item[-] \textbf{Equivalence Query} (hereinafter referred to as \emph{eq}) Given a hypothesis
    (usually constructed by the learning algorithm), \emph{eq} checks whether the hypothesis is
    equivalent to the system-under-learn and generates a counter-example if needed. Generally, 
    \emph{eq} is irrealizable when SUL is a blackbox. So we tactically use $mq$ to achieve the
    approximate results.
\end{itemize}

@@

\subsection{Time Domain}
Time has been investigated in several extensions of Reo. For example, timed
Reo\cite{DBLP:conf/sefm/ArbabBBR04, DBLP:conf/fmoods/MengA07}, hybrid Reo\cite{DBLP:conf/icfem/ChenSS14}, etc.
Generally, these models are designed to handle real-time behavior where time is defined on the
real number field $\mathbb{R}$. However, there are also some works like
\cite{DBLP:journals/fmsd/PrabhakarDM015} where
rational time indeed  makes things easier. In this paper, we choose the rational number field
$\mathbb{Q}$ as our time domain, which simplifies discretization of timed behavior greatly.

As presented in section \ref{sec:reo}, all real-time behavior in timed Reo comes with a finite set
of timed channels. We use
$t_i\in\mathbb{Q}, i=1,2,\dots,n$ to denote the delays of these timed channels, and then we can define a precision
function $prec$ which calculate a gcd-style \emph{maximal time precision}:
\[
prec(t_1,\dots,t_n) = \max\{T\in\mathbb{Q}|\forall t_i.\exists n_i\in\mathbb{N}.t_i=n_i\cdot T\}
\]
It's easy to prove that such a $T$ always exists.

In embedded systems, the maximal time precision is also called ``clock-period'', which is the basic
time unit provided by an oscillator. For example, a widely-used Intel-80C51 microcontroller works
under the frequency 16Mhz, where the clock-period is $62.5$ns. A reasonable assumption comes that
such a maximal time precision $dt=prec(t_1,\dots,t_n)$ should always be provided to make sure
different components are able to work together. Then all $t_i$-Timers can be seen as $n_idt$-Timers
for some $n_i$ and we use $n_i$-Timers instead in the following sections.

@@

\subsection{Parameterized Mealy Machine}
We present a model named \emph{parameterized Mealy machine} to model timed connectors. PMM is
supposed to behave as a bridge between Mealy machines and timed connectors. Connectors are firstly
defined as PMMs, and composed by \emph{product} and \emph{link} operators. Then, concrete
Mealy-machine model can be generated from the PMM model.
Following the formal definition of Mealy machine in Section \ref{sec:preliminaries}, we define the
\emph{parameterized Mealy machine} as follows.

@@

\begin{definition}[Product]
  The product operator \emph{prod} of two PMMs
  \[
  \mathcal{PM'} = prod(\mathcal{PM}_1,\mathcal{PM}_2)\mbox{ if }\mathcal{PM}_2.O\cap
  \mathcal{PM}_1.I=\varnothing
  \]
  is defined as follows.
  \begin{itemize}
  	\item[-] $\forall\Sigma. \mathcal{PM}'.S(\Sigma)=\mathcal{PM}_1.S(\Sigma)\times \mathcal{PM}_2.S(\Sigma)$
    \item[-] $\mathcal{PM}'.I=(\mathcal{PM}_1.I\cup \mathcal{PM}_2.I)\backslash \mathcal{PM}_1.O$
    \item[-] $\mathcal{PM}'.O=(\mathcal{PM}_1.O\cup \mathcal{PM}_2.O)\backslash \mathcal{PM}_2.I$
  \end{itemize}
  \emph{(Here we assume that sink ends of $\mathcal{PM}_1$ can be connected to source ends $\mathcal{PM}_2$, but not
  vise versa)}
  \begin{itemize}
    \item[-] $\mathcal{PM}'.s_0=(\mathcal{PM}_1.s_0, \mathcal{PM}_2.s_0)$
    \item[-] $\forall\Sigma. \mathcal{PM}'.\delta(\Sigma)((s_1,s_2), i)=$
      \begin{displaymath}
        \left\{
        \begin{array}[ht]{ll}
          (Out_1 \cup Out_2)\downarrow_{\mathcal{PM}'.O} & Out_1\neq\rblock\land Out_2\neq\rblock \\
          \rempty & i = T \\
          \rblock & otherwise \\
        \end{array}
        \right.
      \end{displaymath}
      where we have
      \begin{itemize}
        \item[*] $In_1 = i\downarrow_{\mathcal{PM}_1.I}$
        \item[*] $Out_1 = \mathcal{PM}_1.\delta(\Sigma)(s_1,In_1)$
        \item[*] $In_2 = (Out_1 \cup i)\downarrow_{\mathcal{PM}_2.I}$
        \item[*] $Out_2 = \mathcal{PM}_2.\delta(\Sigma)(s_2,In_2)$
      \end{itemize}
    \item[-] $\forall\Sigma. \mathcal{PM}'. \lambda(\Sigma)((s_1,s_2),i)=(s_1',s_2')$
      where we have
      \begin{itemize}
        \item[*] $s_1' = \mathcal{PM}_1.\lambda(\Sigma)(s_1,In_1)$
        \item[*] $s_2' = \mathcal{PM}_2.\lambda(\Sigma)(s_2,In_2)$
      \end{itemize}
  \end{itemize}
\end{definition}

The idea in \emph{production} is quite simple. The output of $\mathcal{PM}_1$ would be provided as part of the
input of $\mathcal{PM}_2$. Time actions T is only executed simultaneously between $\mathcal{PM}_1$ and $\mathcal{PM}_2$.

@@

\subsection{Observation Table}
\emph{Observation Tables}, proposed in \cite{DBLP:journals/iandc/Angluin87}, is used to construct
a hypothetical Mealy machine from the results of membership queries.

@@

From a closed observation we can construct a hypothetical Mealy machine in a straight-forward way.
The Mealy machine $\mathcal{H}=\langle S,s_0,\mathcal{I},\mathcal{O},\delta,\lambda\rangle$ is
constructed from $obs$ as follows:
\begin{itemize}
  \item[-] Every state in $S$ corresponds to an input sequence in $obs.S$
  \item[-] $s_0$ corresponds to the empty sequence $\varepsilon$

    \emph{(we use input sequences to indicate their corresponding states in following parts)}
    \todo{define the empty sequence}
  \item[-] $\delta(s,a)=mq(sa)$
  \item[-] $\lambda(s,a)=acc(sa,obs.D)$
\end{itemize}

@@

Our future works would mainly focus on better support of dense time. To describe dense
time behavior, the Mealy machine model needs a lot of changes instead of a simple T action. Besides,
we will also try to improve our algorithm to handle non-deterministic behaviors. This is not a brand
new topic \cite{DBLP:conf/isola/VolpatoT14}, but we believe that there is still room for
improvement.

\section*{Acknowledgments}
The work was partially supported by the National Natural Science Foundation of China under grant no.
61202069, 61272160 and 61532019, and Research Fund for the Doctoral Program of Higher Education of
China under grant no. 20120001120103.

\bibliographystyle{abbrv}
\bibliography{bib}
